---
title: "Lab 2 - Reading and Exploring Data"
author: "P. Johnson, L. Shaw, B. Rogers"
date: "Wednesday, February 4, 2015"
output: ioslides_presentation

---

## Office hours

- Homework-a-palooza on Mondays 11:30a - 1:30p in Fraser 458
- Tuesday 9:30a - 11:00a Dr. Johnson at the CRMDA
- Tuesday 1:30p - 3:00p Ben's office Blake 311
- Wednesday 10:00a - 11:00a Leslie's office Fraser 10
- Thursday 3:00p - 5:00p Dr. Johnson at the CRMDA
- Friday 1:30p - 3:00p Repeat lab Blake 114

## First homework due Feb. 11 at 5pm

- You can get half credit by attempting to answer each question, whether correct or not.
    + if asked for code, provide code
    + if asked for a table, provide table
    + if asked for plot, provide a plot
    + etc.
- If you minimally answer each question correctly, you can get a B
- You need to demonstrate a deeper understanding in your answer to get full points

## To do
- Accessing internal data sets
- Creating data frames
- Reading and writing data
- Accessing specific entries in data
- Getting to know your data
- Plots

## Internal data sets
```{r eval = FALSE}
install.packages("UsingR") # install UsingR package
```
```{r}
library(UsingR) # load the package to R session
```

## Look at objects in workspace
```{r}
help(package="UsingR") # take a look at what's in the package
ls()   # look at the list of objects in the workspace
```
- Do you see more than just the babies data set?
- Maybe you need to clear your workspace
```{r eval = FALSE}
rm(list = ls())
```

## Looking at whole data set
```{r}
babies  # simply type the name of the data frame
```

## Look at first 6 rows
- Use ?head to find out how to see more or less rows
```{r}
head(babies) # show the first 6 rows
```

## Look at first 6 rows
- Use colnames or rownames function so you know how to refer to your data
```{r}
colnames(babies) # returns column names. Also see rownames()
```

## Creating dataframes
- In lab 1 we saw how to create vectors
- Sometimes we want to put vectors together into a dataframe, a two dimensional object
```{r}
x <- c(100, 95, 93, 97) # define a data vector (variable) x
y <- c(95, 98, 86, 91) # define y
dat <- data.frame(x, y) # create a data frame, cols are variables
## You can see dat in your Workspace now
dat
```

## Working directory
- If you launch your R editor with the R file you are going to use and the data that you will use is in that same directory, then R will treat that folder as your working directory.
- Check your working directory and change if needed
```{r}
getwd()
```
```{r eval = FALSE}
# see lab 1 for examples on setwd()
setwd()
```

## Reading data into R
- There are many ways to read data into R
- We will use read.table() or variations on that function in labs
- All code will assume data is in the same folder as our R file
```{r}
# Check out ?read.table
mydata1 <- read.table(file = "job1.txt")
```

## Reading a comma-delimited file
- Instead of read.table(), we will use read.csv()
- Note the column names that we get by setting header = TRUE
```{r}
mydata <- read.csv(file = "job.csv", header = TRUE)
head(mydata)
```

## Referencing specific entries or rows
```{r}
mydata[1, 2]  # the entry at row 1 and column 2
mydata[1:10, ] # the first ten rows (all columns)
```

## See two columns
```{r}
mydata[ , 4:5] # the 4th and 5th columns
```

## See specific set of rows and columns
```{r}
mydata[c(2, 3, 5), 1:3] # rows 2, 3, and 5 of columns 1-3
```

## See all data except for some rows
```{r}
mydata[-1:-400, ] # all data except for the first 400 rows
```

## Refer to columns by name
```{r}
mydata["1" , "AGE"]
mydata[1:20, c("IQ", "AGE", "WBEING", "SATIS")]
```

## Look at a subset based on a factor variable
```{r}
mydata[mydata$FEMALE == 0, ]  # select data for males
```

## Look at a subset based on a continuous value
```{r}
mydata[mydata$AGE > 50, ]
```

## Reference the IQ column
```{r}
mydata$IQ
```

## Size of data set and basic statistics
```{r}
dim(mydata) # returns dimensions of a data frame
summary(mydata) # a summary of each variable
```

## Look at a summary of 4 columns
```{r}
dim(mydata[ , c("IQ", "AGE", "WBEING", "SATIS")])
summary(mydata[ , c("IQ", "AGE", "WBEING", "SATIS")])
```

## Frequency tables
```{r}
table(mydata$FEMALE, mydata$JOBPERF)
```

## Covariance and correlation
```{r}
cov(mydata$AGE, mydata$JOBPERF)
cor(mydata$AGE, mydata$JOBPERF)
```

## Functions for vectors applied to a dataframe
- All the one-dimensional descriptive functions apply here if we reference one column
```{r}
mean(mydata$IQ)
sd(mydata$IQ)
range(mydata$IQ)
```

## Create variable and add to dataframe
- We can create a new variable in the data frame by directly assigning values to it.
```{r}
mydata$IQ.centered <- mydata$IQ - mean(mydata$IQ)
summary(mydata)
```

## Factor variables in R
When reading in data using data.frame() or read.table()/ read.csv(), R guesses the types of variables. By default, numeric variables will still be numeric, while character data and logical data will be the translated into factors (categorical variables).

But sometimes we need to tell R specifically that some of the numeric variables are actually factors. 

The factor() function converts a numeric variable into a factor. The labels= argument specifies labels for the factor levels. 

## Telling R about a factor
- Create a new variable
- Look at the levels
```{r}
mydata$FEMALE.f <- factor(mydata$FEMALE)
levels(mydata$FEMALE.f) # look at the catagories (aka levels)
```

## Add and change level information
- We can associate labels to numbers in our factor
- Labels need to be assigned in order that R sees the levels (alpha-numeric)
- Note that in the data male = 0 and female = 1
```{r}
mydata$FEMALE.f <- factor(mydata$FEMALE, labels=c("Male", "Female"))
levels(mydata$FEMALE.f)
mydata$TURNOVER.f <- factor(mydata$TURNOVER, labels=c("No", "Yes"))
levels(mydata$TURNOVER.f)
```

## Look at the new summary
```{r}
summary(mydata[ , c("FEMALE", "FEMALE.f", "TURNOVER", "TURNOVER.f")])
```

## Tired of typing mydata$ in front of your variable?
- For functions that don't have 'data =' as an option, use with() instead
```{r}
# Equal to table(mydata$FEMALE, mydata$JOBPERF)
with(mydata, table(FEMALE, JOBPERF))
# Another example
with(mydata, cov(IQ, JOBPERF))
```

## Plots
- Scatterplot
- Boxplot
- Crosstabs
- Bar plots

## Basic scatterplot
```{r}
plot(JOBPERF ~ IQ, data = mydata)
```

## Matrix of scatterplots
- Use all rows of cols 2, 3, 4
```{r}
pairs(mydata[, c(2, 3, 4)]) 
```

## Create new dataframes separated by gender
```{r}
mydata_m <- mydata[mydata$FEMALE==0, ]  # copy rows where FEMALE==0
mydata_f <- mydata[mydata$FEMALE==1, ]  # copy rows where FEMALE==1
```

## Redraw scatterplot frame and then add points for males
```{r}
plot(JOBPERF ~ IQ, data = mydata, xlab="IQ", ylab="Job Performance", 
     type="n")
points(JOBPERF ~ IQ, data = mydata_m, col="black", 
       pch=1) #  col= sets the color
```

## Syntax to add points for female and legend
```{r eval = FALSE}
plot(JOBPERF ~ IQ, data = mydata, xlab="IQ", ylab="Job Performance", 
     type="n")
points(JOBPERF ~ IQ, data = mydata_m, col="black", 
       pch=1) #  col= sets the color
points(JOBPERF ~ IQ, data = mydata_f, col="grey80", 
       pch=4)  # pch= sets the shape
legend("topleft", legend=c("Male","Female"), col=c("black","grey80"), 
       pch=c(1,4))
```

## Syntax to add points for female and legend
```{r echo = FALSE}
plot(JOBPERF ~ IQ, data = mydata, xlab="IQ", ylab="Job Performance", type="n")
points(JOBPERF ~ IQ, data = mydata_m, col="black", pch=1) #  col= sets the color
points(JOBPERF ~ IQ, data = mydata_f, col="grey80", pch=4)  # pch= sets the shape
legend("topleft", legend=c("Male","Female"), col=c("black","grey80"), pch=c(1,4))
```

## Boxplot of data with outliers
```{r}
boxplot(mydata$IQ)
```

## And another boxplot
```{r}
boxplot(mydata$JOBPERF)
```

## Boxplot of a subgroup with labels
```{r eval = FALSE}
boxplot(mydata$JOBPERF ~ mydata$FEMALE.f)  # accepts an equation
boxplot(mydata$JOBPERF ~ mydata$FEMALE.f, ylab="Job Performance", 
        xlab="Gender", names=c("Male","Female"))  #  Add labels 

# Equivalent to first plot above but cleaner to read
boxplot(JOBPERF ~ FEMALE.f, data=mydata)

# Commented out in lab code but try running it!
```

## Cross-Tabulation
```{r eval = FALSE}
# Run this line if you need to install the package
install.packages("descr")
```
```{r}
library(descr)
myTable1 <- with(mydata, CrossTable(FEMALE.f, TURNOVER.f))
```

## Chi-Square output
```{r}
myTable1
```

## Contingency table function
```{r eval = FALSE}
# Run this line if you need to install the package
install.packages("memisc")
```
```{r}
library(memisc)
```

## Contingency table with counts and percents
```{r}
myTable2 <- with(mydata, genTable(TURNOVER.f ~ FEMALE.f))
myTable2
myTable3 <- with(mydata, genTable(percent(TURNOVER.f) ~ FEMALE.f))
myTable3
```

## Barplot
- A bar plot is a graphic presentation of a cross tabulation table.
```{r eval = FALSE}
barplot(myTable3[1:2,], beside=T, names=c("Men","Women"), 
        col=c("blue","red"), density=c(90,40), angle=c(45,-45),
        ylab="Turnover", xlab="Gender")
legend("topright", legend = levels(mydata$TURNOVER.f), 
       fill=c("blue","red"), density=c(90,40), angle = c(45,-45),
       cex=1.2)

# Note that barplot() uses "col=" but legend() uses "fill="
```

## Barplot
- A bar plot is a graphic presentation of a cross tabulation table.
```{r echo = FALSE}
barplot(myTable3[1:2,], beside=T, names=c("Men","Women"), 
        col=c("blue","red"), density=c(90,40), angle=c(45,-45),
        ylab="Turnover", xlab="Gender")
legend("topright", legend = levels(mydata$TURNOVER.f), 
       fill=c("blue","red"), density=c(90,40), angle = c(45,-45),
       cex=1.2)
```



## Exporting data
- Like read.table or read.csv there are corresponding write functions
- Double check your working directory using these functions
```{r eval = FALSE}
write.table(mydata, file="out.txt", sep="\t", row.names=FALSE, 
            quote = FALSE)
write.csv(mydata, file="out.csv", row.names=FALSE, quote = FALSE)
```
- 'sep =' is used to denote how columns should be separated
- tab is used above; comma ',' and space ' ' are also common