---
title: "Getting Started With R"
author: "P. Johnson, B. Rogers, and L. Shaw"
date: "Wednesday, January 28, 2015"
output: slidy_presentation
---

## Getting comfortable with R

Today's lab will attempt to introduce you to R, teach you some basics about how R is set up, how to create some data, how to run some basic functions, and how to look for help. 

If you do not find what you need through R help commands, search the internet. There are many examples, and if you do not understand one example, keep searching. Someone will write about that function in a way that will make sense to you.

## More than just a calculator

- Interacting with R
- Using R as a calculator
- Assigning information to a variable
- Getting help

## Working with data

- Using R to create a vector
- Descriptive statistics
- Vector arithmetic
- Referencing specific data
- Plots
- Your workspace
- Packages and __rockchalk__

## Interacting with R
- Enter a command in the console, or
- Use one of the editors and send to the console
- Liberally comment your code and save
- Find code that works and modify for your own use

## The R console
- Enter code below in a console
- Enter code in editor and send to R console 
```{r}
# Comment to describe code
# What summary provides depends on what is in the parentheses
# cars is a built in data set so you get summary statistics
summary(cars)
```

## Basic arithmetic

```{r}
3 + 2  # Addition
3 - 1  # Subtraction
7 * 8  # Multiplication
9 / 3  # Division
```

## More advanced calculations

```{r}
3 ^ 2  # Power
sqrt(36) # Square root
log(7) # Natural logarithm ln
```

## Using R as a calculator
```{r}
exp(10) # e^10
```
- Use parentheses around functions to force order of operations
```{r}
1 + 3 * log(10) - (exp(7) - 3) / 4
```

## Assigning information to a variable
- You can store single values, vectors, matrices, data frames, arrays, and other objects
- The matrix object can only hold numbers
- Numbers or strings can be stored in the other object types
- Good form uses __<-__ to assign something to a variable
- X and x are two different variables because R is case-sensitive

## Assigning numbers to a variable
```{r}
x <- 1
x   # The value can be printed by typing the object name
x <- 2 + 5  # The result is saved in the object 'x'
x
```

## Assigning text to a variable
```{r}
z <- "Hello!"  # We may save an object as text.
z
```

## Assigning numbers to X and x
```{r}
X <- 5
X + x
X <- log(2) # This will overwrite the previous value.
X + x
```

## Getting help
```{r eval = FALSE}
## When R was installed, HTML format help files were copied 
## onto your hard drive. To access these files, just type
help.start()
## To request an R document for a specific function, use '?'. 
?log
## To request help by keywords, use '??'.
??logarithm
```

## Store data in a vector
- Many values can be stored together in vectors (c is combine)
```{r}
midterm <- c(99, 87, 96, 100, 82, 79, 88, 85, 94, 90)
midterm
f <- 1:20
f
g <- seq(from = 1, to = 20) # seq(start value, end value, interval)
h <- seq(1, 20, by = 0.2) # specify interval
```

## More functions to create vectors
- rep stands for replicate
```{r}
k <- rep(1, 5)  # rep(value, number of times)
k
l <- rep(1:10, 5)  # rep() can accept numbers, text, or NA
l
```

## Generating random values
- What is required in the function depends on the type of distribution
```{r}
datn <- rnorm(n = 1000, mean = 24, sd = 3)
# Look at datn to see what R generated

datp <- rpois(1000, 3)     # 1000 random numbers 
## from a poisson disribution
## with an expected value of 3. 
```

## Descriptive statistics
- Re-create midterm variable
- Compute the mean
```{r}
midterm <- c(99, 87, 96, 100, 82, 79, 88, 85, 94, 90)
mean(midterm) 
```
- Run one of the following functions
- median, var, sd, min, max, range, length, table, sum, summary

## What if you are missing data?
```{r}
midterm2 <- c(99, NA, 96, 100, 82, NA, 88, 85, 94, 90)
mean(midterm2)
```
- Tell the function to compute even though there is missing data
```{r}
mean(midterm2, na.rm = TRUE)
```

## Simple vector arithmetic
- R can save you time if you understand how whole vectors can be processed
```{r}
c(1, 2, 3, 4) - 4
c(1, 2, 3, 4)/4
c(1, 2, 3, 4)/c(4, 3, 2, 1)   # 1/4, 2/3, 3/2, 4/1
```

## Vector arithmetic with functions
- R can save you time if you understand how whole vectors can be processed
```{r}
log(c(1, 2, 3, 4))  # applies log to each element
## Can you tell what this is?
sum((midterm - mean(midterm))^2)/(length(midterm) - 1)
```

## Referencing specific data in vector
- You can use individual values in a vector or other object
```{r}
midterm
midterm[2]
midterm[length(midterm)] # What is happening here?
```

## Reference more than one thing
```{r}
midterm[1:4] # take more than one element at a time.
midterm[c(1:4, 8, 10)]
```
```{r eval = FALSE}
midterm[1, 2, 3, 4, 8, 10]  # This is NOT THE SAME
## R could only evaluate this if midterm was an array with
## 6 dimensions.
```

## Histogram help
```{r}
mydata <- rnorm(1000, 0, 1) 

?hist
```


## Histogram with frequencies
```{r}
hist(mydata)
```

## Histogram with proportions
- Values on y-axis are sometimes referred to as the following: proportion, probability, and density
```{r}
hist(mydata, prob = TRUE) 
```

## Histogram with labels
```{r}
hist(mydata, prob=TRUE, xlab="X-Variable Name", 
     ylab="Proportion of Cases") # Label x and y axes
```

## Add a title and number of bins
```{r}
hist(mydata, prob=TRUE, xlab="X-Variable Name", 
     ylab="Proportion of Cases", main="My First Histogram", breaks=50) 
```

## Add a density line
```{r tidy=TRUE}
hist(mydata, prob=TRUE, xlab="X-Variable Name", 
     ylab="Proportion of Cases", breaks=50) 
lines(density(mydata))
```

## Your working directory
- When reading data or saving objects to your computer, you need to know where R is looking
```{r}
getwd()
```
- To change where R is looking, specify relative or full path
```{r eval = FALSE}
setwd("Labs")
setwd("D:\\Users\\l076s857\\Desktop") # For Windows
setwd("D:/Users/l076s857/Desktop") # For Windows, Mac, Linux
```
## Your workspace
- List all items
```{r}
ls()
```
- Remove one item, then all
```{r}
rm(x)
rm(list = ls(all = TRUE))
```

## Information stored in objects
```{r}
mydata <- rnorm(1000, 0, 1) 
h1 <- hist(mydata, prob=TRUE, xlab="X-Variable", ylab="Density", 
           main="My First Histogram", breaks=10) 
```

## Information stored in histograms {.allowframebreaks}
```{r}
options(width=55)
h1
```

## Information stored in objects
```{r}
attributes(h1) 
names(h1)
h1$mids   # The $ allows you to access a specific attribute of the object
```

## Working with packages
- Some functions are built into R, like all of the functions that we have used so far.
- There are many functions in other packages and a list of packages can be found at http://cran.r-project.org/web/packages/available_packages_by_name.html
- The first time you want to use a function in a new package, you need to install it.
- Once the package is installed, you will need to reference the package in your R code before trying to use that function.

## Install and use package
- If you do not have admin rights, a local directory will be created by R for the packages you install.
```{r eval = FALSE}
install.packages("rockchalk", dep = TRUE)
```
```{r}
library(rockchalk)
```

## Use function from rockchalk
```{r}
mydata <- rnorm(1000, 0, 1) 
summarize(mydata)
```
